A Standardized Effect Size for Evaluating the Comparing the Strength of Phylogenetic Signal

### 2020-07-16 Manuscript/OldVsns/Manuscript-4.tex

@@

Evolutionary biologists commonly seek to describe the relative strength
of phylogenetic signal in phenotypic traits, to determine the extent to
which shared evolutionary history has influenced trait covariation among
taxa. This is often accomplished by interpreting empirical estimates of
\(\lambda\); with smaller values signifying `weak' phylogenetic signal,
while larger values are interpreted as `strong' phylogenetic signal
(e.g., De Meester et al. 2019; Pintanel et al. 2019; Su et al. 2019).
Other approaches for interpreting \(\lambda\) are more statistical. For
instance, some have evaluated whether the observed \(\lambda\) differs
from some expected value through the use of confidence intervals
(Vandelook et al. 2019) or by performing likelihood ratio tests that
compare the observed model fit to that obtained when \(\lambda=0\) or
\(\lambda=1\) (Freckleton et al. 2002; Cooper et al. 2010; Bose et al.
2019). Additionally, qualitative comparisons of \(\lambda\) estimates
obtained from multiple phenotypic traits have been used to infer whether
the strength of phylogenetic signal is greater in one trait as compared
to another (e.g., Liu et al. 2019; Bai et al. 2019). Indeed, statements
regarding the strength of phylogenetic signal based on \(\lambda\) are
rather common in the evolutionary literature. For instance, of the 204
papers published in 2019 that estimated and reported Pagel's \(\lambda\)
(found in Google.scholar), 40\% interpreted the strength of phylogenetic
signal for at least one phenotypic trait. Additionally, because nearly
half of the 1,572 \(\lambda\) values reported were near the limits of
the parameter (Figure 1), this percentage is even higher, as biological
interpretation of phylogenetic signal at the limits of \(\lambda\) are
known. \hfill\break

@@

It seems intuitive to interpret the strength of phylogenetic signal
based on the value of \(\lambda\), as \(\lambda\) is a parameter on a
bounded scale (\(0\to1\)) for which interpretation of its extremal
points are understood. Specifically, \(\lambda=0\) represents no
phylogenetic signal, while \(\lambda=1\) is phylogenetic signal as
expected under Brownian motion. However, equating values of \(\lambda\)
directly to the strength of phylogenetic signal presumes two important
statistical properties that have not been fully explored. First, it
presumes that values of \(\lambda\) can be precisely estimated, as
biological inferences regarding the strength of phylogenetic signal
depend on high accuracy in its estimation. Therefore, understanding the
precision in estimating \(\lambda\) is paramount. One study (Boettiger
et al. 2012) found that estimates of Pagel's \(\lambda\) displayed less
variation (i.e., greater precision) when data were simulated on a large
phylogeny (\(N=281\)) as compared to a small one (\(N=13\)). From this
observation it was concluded that insufficient data (i.e., the number of
species) was the underlying cause of the increased variation across
parameter estimates (Boettiger et al. 2012). Indeed, such a pattern is
common with statistical estimators, as summary statistics and parameters
are often more precise at greater sample sizes (Cohen 1988). However,
this conclusion also assumes that the precision of \(\lambda\) remains
constant across its range (\(\lambda = 0 \to 1\)); an assumption that to
date, has not been verified. Thus, despite widespread use of Pagel's
(1999) \(\lambda\) in macroevolutionary studies, at present, we still
lack a general understanding of the precision with which \(\lambda\) can
estimate levels of phylogenetic signal in phenotypic datasets.
\hfill\break

Second, while estimates of \(\lambda\) are within a bounded scale
(\(0\to1\)), this does not \emph{de-facto} imply that the estimated
values of this parameter correspond to the actual strength of the
underlying input signal in the data. For this to be the case,
\(\lambda\) must be a statistical effect size. Effect sizes are a
measure the magnitude of a statistical effect in data, represented on a
common scale (Glass 1976; Cohen 1988). Effect sizes have widespread use
in many areas of the quantiative sciences, as they represent measures
that may be readily summarized across datasets as in meta-analysis
(Glass 1976; Hedges and Olkin 1985; Arnqvist and Wooster 1995), or
compared among datasets (e.g., Adams and Collyer 2016, 2019a).
Unfortunatley, not all model parameters and test statistics are effect
sizes, and thus many summary measures must first be converted to
standardized units (i.e., an effect size) for meaningful comparison (see
Rosenthal 1994). As a consequence, it follows that only if \(\lambda\)
is a statistical effect size can comparisons of estimates across
datasets be interpretable. For the case of \(\lambda\), this has not yet
been explored. \hfill\break

In this study, we evaluate the precision of Pagel's \(\lambda\) for
estimating known levels of phylogenetic signal in phenotypic data. We
use computer simulations with differing numbers of species, differently
shaped phylogenies, and differing input levels of phylogenetic signal,
to explore the degree to which \(\lambda\) correctly identifies known
levels of phylogenetic signal, and under what circumstances. We find
that estimates of \(\lambda\) vary widely for a given input value of
phylogenetic signal, and that the precision in estimating \(\lambda\) is
not constant across its range. Rather, there is decreased precision when
input levels of phylogenetic signal are of intermediate strength.
Additionally, the same estimated values of \(\lambda\) may be obtained
from datasets containing vastly different input levels of phylogenetic
signal. Thus, \(\lambda\) is not a reliable estimate of the strength of
phylogenetic signal in phenotypic data. We then describe a standardized
effect size for measuring the strength of phylogenetic signal in
phenotypic datasets, and apply the concept to two common measures of
phylogenetic signal: \(\lambda\) and \(Kappa\). Through simulations
across a wide range of conditions, we find that the precision of effect
sizes based on \(\lambda\) (\(Z_{\lambda}\)) are less reliable than that
those based on \(Kappa\) (\(Z_K\)), implying that \(Z_K\) is a more
robust effect size measure. We also propose a two-sample test statistic
that may be used to compare the strength of phylogenetic signal among
datasets, and provide an empirical example to demonstrate its use. We
conclude that estimates of phylogenetic signal using Pagel's \(\lambda\)
are often inaccurate, and thus interpreting strength of phylogenetic
signal in phenotypic datasets based on this measure is compromised. By
contrast, effect sizes obtained from \(Kappa\) hold promise for
characterizing phylogenetic signal, and for comparing the strength of
phylogenetic signal across datasets.

@@

We conducted a series of computer simulations to evaluate the precision
of Pagel's \(\lambda\). Our primary simulations were based on pure-birth
phylogenies; however, we also evaluated patterns on both balanced and
pectinate trees to determine whether tree shape affected our findings
(see Supporting Information). First we generated 50 pure-birth
phylogenies at each of six different tree sizes, ranging from 32 to 1024
taxa (\(n=2^5 - 2^{10}\)). Next, we rescaled the simulated phylogenies
by multiplying the internal branches by \(\lambda_{in}\), using 21
intervals of 0.05 units across its range
(\(\lambda_{in} = 0.0 \to 1.0\)), resulting in 1050 scaled phylogenies
at each level of species richness (\(n\)). Continuous traits were then
simulated on each phylogeny under a Brownian motion model of evolution
to obtain datasets with differing levels of phylogenetic signal, that
ranged from no phylogenetic signal (when \(\lambda_{in} =0\)), to
phylogenetic signal corresponding reflecting Brownian motion (when
\(\lambda_{in} =1\)). For each dataset we then estimated phylogenetic
signal (\(\lambda_{est}\)), and calculated the precision of \(\lambda\)
using the variance (\(\sigma^2_\lambda\)) across datasets at each input
level of phylogenetic signal and level of species richness. \hfill\break

We also evaluated the precision of \(\lambda\) when estimated in PGLS
regression and ANOVA (i.e., \(Y\sim{X}\)). Here, an independent variable
\(X\) was simulated on each phylogeny under a Brownian motion model of
evolution (for PGLS regression). For phylogenetic ANOVA, random groups
(\(X\)) were obtained by simulating a discrete (binary) character on
each phylogeny. Next, the dependent variable was simulated in such a
manner as to contain a known relationship with \(X\) plus random error
containing phylogenetic signal. This was accomplished as:
\(Y=\beta{X}+\epsilon\). Here, the association between \(Y\) and \(X\)
was modeled using a range of values:
\(\beta=(0.0,0.25, 0.5, 0.75,1.0)\), and the residual error was modeled
to contain phylogenetic signal simulated under a Brownian motion model
of evolution: \(\epsilon=\mathcal{N}(\mu=0,\sigma=\mathbf{C})\): (see
Revell 2010 for a similar simulation design). The fit of the
phylogenetic regression was estimated using maximum likelihood, and
parameter estimates (\(\beta_{est}\) and \(\lambda_{est}\)) were
obtained. Precision estimates (\(\sigma^2_\lambda\)) at each input level
of phylogenetic signal and level of species richness were then observed.
\hfill\break

@@

\emph{Results.} We found that the precision of \(\lambda_{est}\) varied
widely across simulation conditions. Predictably, precision improved as
the number of species increased (Figure 2). This confirmed earlier
findings of Boettiger et al.~(2012), and adhered to parametric
statistical theory. However, in many cases the set of \(\lambda_{est}\)
spanned nearly the entire range of possible values (e.g., \(n=32\);
\(\lambda_{in}=0.5\): \(\lambda_{est}= 0.0\to 0.985\)), revealing that
estimates of \(\lambda\) were not a reliable indicator of input
phylogenetic signal. Importantly, the precision of \(\lambda_{est}\) was
not uniform across all levels of phylogenetic signal. The worst
precision was observed at intermediate levels of phylogenetic signal
(\(\lambda_{in}\approx0.5\)), while precision improved as input levels
approached the extremes of \(\lambda\)'s range (i.e.,
\(\lambda_{in}\rightarrow0\) \& \(\lambda_{in}\rightarrow1\)). Thus,
estimates of \(\lambda\) were least reflective of the true input signal
at intermediate values. Additionally, even at large levels of species
richness, we found that the range of \(\lambda_{est}\) still encompassed
a substantial portion of possible values (e.g., \(n=512\);
\(\lambda_{in}=0.5\): \(\lambda_{est} = 0.32\to 0.68\)). Likewise, the
same \(\lambda_{est}\) could be obtained from datasets containing vastly
different input levels of phylogenetic signal (e.g., \(n=512\);
\(\lambda_{est} = 0.5\); \(\lambda_{in} = 0.25 \to 0.65\)). These
findings were particularly unsettling when considered in light of our
literature survey. Over one quarter of the \(\lambda\) estimates
obtained in empirical studies (421 of 1,572) were between
\(\lambda=0.25\) and \(\lambda=0.75\) (Figure 1). This range reflected
the region that our simulations identified as being the least reliable
in terms of accurately characterizing levels of phylogenetic signal, yet
30\% of these empirical estimates were explicitly interpreted in terms
of the strength of phylogenetic signal that they represented (i.e.,
weak, intermediate, strong phylogenetic signal). \hfill\break

Finally, when \(\lambda\) was co-estimated with regression parameters in
PGLS regression, the results of our simulations were quite similar.
Here, regression parameters (\(\beta\)) were accurately estimated,
confirming earlier findings of Revell (2010) (see Supporting
Information). However, estimates of phylogenetic signal (\(\lambda\))
were less precise (Figure 3), and the spread of \(\lambda_{est}\) was
similar to that observed when \(\lambda\) was estimated for only the
dependent variable, as in Figure 2. Taken together, these findings
reveal that \(\lambda_{est}\) does not precisely characterize observed
levels of phylogenetic signal in phenotypic datasets, and that
biological interpretations of the strength of phylogenetic signal based
on \(\lambda\) may be highly inaccurate. \hfill\break

@@

Adams and Collyer (2019b) suggested that the strength of phylogenetic
signal could be represented as an effect size, based on the \(Kappa\)
statistic and its empirical sampling distribution from permutation. Here
we formalize that suggestion, resulting in an effect size of:

\begin{align}
    Z_{K}=\frac{K_{obs}-\hat\mu_K}{\hat\sigma_K}
\end{align}

where \(K_{obs}\) is the observed phylogenetic signal, and \(\hat\mu_K\)
and \(\hat\sigma_K\) are the mean and standard deviation of the
empirical sampling distribution of \(Kappa\) obtained via permutation.
Similarly, an effect size based on \(\lambda\) could be envisioned as:

@@

In this case, \(\lambda_{obs}\) and \(\hat\sigma_\lambda\) are
empirically derived using maximum likelihood. Note that under the null
hypothesis, no phylogenetic signal is expected (Freckleton et al. 2002),
and thus \(E(\lambda)=0\) under this condition. \hfill\break

To evaluate the utility of \(Z_K\) and \(Z_\lambda\) we calculated both
effect sizes for the simulated datasets generated above, and summarized
the precision of each using its variance (\(\sigma^2_{Z_K}\) and
\(\sigma^2_{Z_\lambda}\)). Results are found in Figure 4 (additional
results are found in the Supporting Information). Here two things are
evident. First, estimates of \(Z_K\) track the input phylogenetic signal
in a more linear fashion than do estimates of \(Z_\lambda\) (Figure
4A,B). Thus, actual changes in the strength of phylogenetic signal are
reflected more evenly in the corresponding values of the effect size
\(Z_K\). Second, the precision of \(Z_K\) is considerably more stable as
compared with \(Z_\lambda\). This may be seen by calculating the
coefficients of variation for the set of precision estimates (i.e.,
\(\sigma^2_{Z_K}\) and \(\sigma^2_{Z_\lambda}\)) across input levels of
phylogenetic signal. Here coefficients of variation in the precision of
\(Z_K\) were an order of magnitude smaller for than for \(Z_\lambda\)
(Figure 4C,D). This implied that estimates of the strength of
phylogenetic signal were more reliable and robust when using \(Z_K\) as
compared with \(Z_\lambda\). \hfill\break

@@

Once the magnitude of phylogenetic signal is characterized using
\(Z_K\), one may wish to compare such measures across datasets, to
determine whether the strength of phylogenetic signal is significantly
greater in one phenotypic trait as compared with another. As with other
effect sizes derived from permutation distributions (e.g., Adams and
Collyer 2016, 2019a), a two-sample test statistic may be found as:

@@

One important advantage of the approach advocated here is that the
resulting effect sizes (\(Z_K\)) are dimensionless, as the units of
measurement cancel out during the calculation of \(Z\) (Sokal and Rohlf
2012). As a consequence, the effect sizes represent the strength of
phylogenetic signal on a common and comparable scale, measured in
standard deviation units, regardless of the initial units and scale of
the original phenotypic variables under investigation. This means that
conceivably, the strength of phylogenetic signal may be compared across
datsets for phenotypic traits in different units and scale (e.g.,
behavioral and morphological traits), so long as they are continuous
variables. Note that this is directly related to the common practice of
converting data to their standard normal deviates prior to their use in
multivariate analysis, if the original variables are expressed in
different units or scales (see Everitt and Hothorn 2011; Legendre and
Legendre 2012; Adams and Collyer 2019b).

@@

\hypertarget{conclusions-and-implications}{%
\section{Conclusions and
Implications}\label{conclusions-and-implications}}

1: summary paragraph

2: expand on Lambda.. lambda innacurate, not precise, level of precision
varies with input physig (worse in mid-range). NEW RESULT. We are first
to show this. NOTE: pattern is obvious with reflection. Since it is a
`bounded' parameter estimation should be best at the extremes\ldots{}
(state this?).. hmm.

Patterns worse with PGLS, though beta still estimated properly.
Conclusion, lambda not overly useful.

3: By contrast, effect size Z-K useful, equally precise across range of
values. Can be used to characterize the strength of physignal, and
because robust to input levels, etc. may be used to compare across
datasets.

Somewhere, recognize that this is somewhat `backwards' from prior
recommendations where Kappa had somewhat lower performance in terms of
type I and type II error (which?? I forget). However, recall that those
studies did not examine the precision of the estimates. Nor was Z-k
included, because it was not yet invented. So Use of Z-k should make
good sense here.

Closing paragraph.

More discussion paragraphs

>>>>

Macroevolutionary studies frequently characterize the phylogenetic
signal in phenotypes, and wish to compare the strength of that signal
across traits. However, analytical tools for such comparisons have
largely remained underdeveloped. In this study, we evaluated the
efficacy of one commonly used parameter (Pagel's \(\lambda\)) to
estimate the strength of phylogenetic signal in phenotypic traits, and
evaluate the degree to which \(\lambda\) correctly identifies known
levels of phylogenetic signal. We find that the precision of \(\lambda\)
in estimating actual levels of phylogenetic signal is often inaccurate,
and that biological interpretations of the strength of phylogenetic
signal based on \(\lambda\) are therefore compromised. We then propose a
standardized effect size based on \(Kappa\) (\(Z_K\)), which measures
the strength of phylogenetic signal, and places it on a common scale for
statistical comparison. Tests based on \(Z_K\) provide a mechanism for
formally comparing the strength of phylogenetic signal across datasets,
in much the same manner as effect sizes may be used to summarize
patterns in quantitative meta-analysis. Our approach extends the
phylogenetic comparative toolkit to address hypotheses that compare the
strength of phylogenetic signal various phenotypic traits, even when
those traits are found in different evolutionary lineages or of in
different units or scale.

@@

Evolutionary biologists commonly seek to describe the relative strength
of phylogenetic signal in phenotypic traits, to determine the extent to
which shared evolutionary history has influenced trait covariation among
taxa. This is often accomplished by interpreting empirical estimates of
\(\lambda\); with smaller values signifying `weak' phylogenetic signal,
while larger values are interpreted as `strong' phylogenetic signal
(e.g., De Meester et al. 2019; Pintanel et al. 2019; Su et al. 2019).
Other approaches for interpreting \(\lambda\) are more statistical. For
instance, some have evaluated whether the observed \(\lambda\) differs
from some expected value through the use of confidence intervals
(Vandelook et al. 2019) or by performing likelihood ratio tests that
compare the observed model fit to that obtained when \(\lambda=0\) or
\(\lambda=1\) (Freckleton et al. 2002; Cooper et al. 2010; Bose et al.
2019). Additionally, qualitative comparisons of \(\lambda\) estimates
obtained from multiple phenotypic traits have been used to infer whether
the strength of phylogenetic signal is greater in one trait as compared
to another (e.g., Liu et al. 2019; Bai et al. 2019). Indeed, statements
regarding the strength of phylogenetic signal based on \(\lambda\) are
rather common in the evolutionary literature. For instance, of the 204
papers published in 2019 that estimated and reported Pagel's \(\lambda\)
(found in Google.scholar), 40\% interpreted the strength of phylogenetic
signal for at least one phenotypic trait. Further, because nearly half
of the 1,572 \(\lambda\) values reported were near 0 or 1 (Figure 1)
where the biological interpretation of \(\lambda\) is known, this
percentage is even higher.

@@

It seems intuitive to interpret the strength of phylogenetic signal
based on the value of \(\lambda\), as \(\lambda\) is a parameter on a
bounded scale (\(0\to1\)) for which interpretation of its extremal
points are understood. Specifically, \(\lambda=0\) represents no
phylogenetic signal, while \(\lambda=1\) is phylogenetic signal as
expected under Brownian motion. However, equating values of \(\lambda\)
directly to the strength of phylogenetic signal presumes two important
statistical properties that have not been fully explored. First, it
presumes that values of \(\lambda\) can be precisely estimated, as
biological inferences regarding the strength of phylogenetic signal
depend on high accuracy in its estimation. Therefore, understanding the
precision in estimating \(\lambda\) is paramount. One study (Boettiger
et al. 2012) found that estimates of Pagel's \(\lambda\) displayed less
variation (i.e., greater precision) when data were simulated on a large
phylogeny (\(N=281\)) as compared to a small one (\(N=13\)). From this
observation it was concluded that insufficient data (i.e., the number of
species) was the underlying cause of the increased variation across
parameter estimates (Boettiger et al. 2012). Indeed, such a pattern is
common with statistical estimators, as summary statistics and parameters
are often more precise at greater sample sizes (Cohen 1988). However,
this conclusion also implies that the precision of \(\lambda\) remains
constant across its range (\(\lambda = 0 \to 1\)); an assumption that to
date, has not been verified. Thus, despite widespread use of Pagel's
(1999) \(\lambda\) in macroevolutionary studies, at present, we lack a
general understanding of the precision with which \(\lambda\) can
estimate levels of phylogenetic signal in phenotypic datasets.
\hfill\break

Second, while estimates of \(\lambda\) are within a bounded scale
(\(0\to1\)), this does not \emph{de-facto} imply that the estimated
values of this parameter correspond to the actual strength of the
underlying input signal in the data. For this to be the case,
\(\lambda\) must be a statistical effect size. Effect sizes are a
measure the magnitude of a statistical effect in data, represented on a
common scale (Glass 1976; Cohen 1988). Effect sizes have widespread use
in many areas of the quantiative sciences, as they represent measures
that may be readily summarized across datasets as in meta-analyses
(Glass 1976; Hedges and Olkin 1985; Arnqvist and Wooster 1995), or
compared among datasets (e.g., Adams and Collyer 2016, 2019a).
Unfortunatley, not all model parameters and test statistics are effect
sizes, and thus many summary measures must first be converted to
standardized units (i.e., an effect size) for meaningful comparison (see
Rosenthal 1994). As a consequence, it follows that only if \(\lambda\)
is a statistical effect size can comparisons of estimates across
datasets be interpretable. For the case of \(\lambda\), this has not yet
been explored. \hfill\break

In this study, we evaluate the precision of Pagel's \(\lambda\) for
estimating known levels of phylogenetic signal in phenotypic data. We
use computer simulations with differing numbers of species, differently
shaped phylogenies, and differing input levels of phylogenetic signal,
to explore the degree to which \(\lambda\) correctly identifies known
levels of phylogenetic signal, and under what circumstances. We find
that estimates of \(\lambda\) vary widely for a given input value of
phylogenetic signal, and that the precision in estimating \(\lambda\) is
not constant across its range. Rather, there is decreased precision when
input levels of phylogenetic signal are of intermediate strength.
Additionally, the same estimated values of \(\lambda\) may be obtained
from datasets containing vastly different input levels of phylogenetic
signal. Thus, \(\lambda\) is not a reliable indicator of the strength of
phylogenetic signal in phenotypic data. We then describe a standardized
effect size for measuring the strength of phylogenetic signal in
phenotypic datasets, and apply the concept to two common measures of
phylogenetic signal: \(\lambda\) and \(Kappa\). Through simulations we
find that the precision of effect sizes based on \(\lambda\)
(\(Z_{\lambda}\)) are less reliable than that those based on \(Kappa\)
(\(Z_K\)), implying that \(Z_K\) is a more robust effect size measure.
We also propose a two-sample test statistic that may be used to compare
the strength of phylogenetic signal among datasets, and provide an
empirical example to demonstrate its use. We conclude that estimates of
phylogenetic signal using Pagel's \(\lambda\) are often inaccurate, and
thus interpreting strength of phylogenetic signal in phenotypic datasets
based on this measure is compromised. By contrast, effect sizes obtained
from \(Kappa\) hold promise for characterizing phylogenetic signal, and
for comparing the strength of phylogenetic signal across datasets.

@@

We conducted a series of computer simulations to evaluate the precision
of Pagel's \(\lambda\). Our primary simulations were based on pure-birth
phylogenies; however, we also evaluated patterns on both balanced and
pectinate trees to determine whether tree shape affected our findings
(see Supporting Information). First we generated 50 pure-birth
phylogenies at each of six different tree sizes, ranging from 32 to 1024
taxa (\(n=2^5 - 2^{10}\)). Next, we rescaled the simulated phylogenies
by multiplying the internal branches by \(\lambda_{in}\), using 21
intervals of 0.05 units across its range
(\(\lambda_{in} = 0.0 \to 1.0\)), resulting in 1050 scaled phylogenies
at each level of species richness (\(n\)). Continuous traits were then
simulated on each phylogeny under a Brownian motion model of evolution
to obtain datasets with differing levels of phylogenetic signal, that
ranged from no phylogenetic signal (when \(\lambda_{in} =0\)), to
phylogenetic signal reflecting Brownian motion (when
\(\lambda_{in} =1\)). For each dataset we then estimated phylogenetic
signal (\(\lambda_{est}\)), and calculated the variance of \(\lambda\)
(\(\sigma^2_\lambda\)) across datasets at each input level of
phylogenetic signal and level of species richness as an estimate of
precision. \hfill\break

We also evaluated the precision of \(\lambda\) when estimated in PGLS
regression and ANOVA (i.e., \(Y\sim{X}\)). Here, an independent variable
\(X\) was simulated on each rescaled phylogeny under a Brownian motion
model of evolution (for PGLS regression). For phylogenetic ANOVA, random
groups (\(X\)) were obtained by simulating a discrete (binary) character
on each phylogeny. Next, the dependent variable was simulated in such a
manner as to contain a known relationship with \(X\) plus random error
containing phylogenetic signal. This was accomplished as:
\(Y=\beta{X}+\epsilon\). Here, the association between \(Y\) and \(X\)
was modeled using a range of values:
\(\beta=(0.0,0.25, 0.5, 0.75,1.0)\), and the residual error was modeled
to contain phylogenetic signal simulated under a Brownian motion model
of evolution on each rescaled phylogeny:
\(\epsilon=\mathcal{N}(\mu=0,\sigma=\sigma^2\mathbf{C})\): (see Revell
2010 for a similar simulation design). The fit of the phylogenetic
regression was estimated using maximum likelihood, and parameter
estimates (\(\beta_{est}\) and \(\lambda_{est}\)) were obtained. We then
calculated precision estimates (\(\sigma^2_\lambda\)) at each input
level of phylogenetic signal and level of species richness. \hfill\break

@@

\emph{Results.} We found that the precision of \(\lambda_{est}\) varied
widely across simulation conditions. Predictably, precision improved as
the number of species increased (Figure 2). This confirmed earlier
findings of Boettiger et al.~(2012), and adhered to parametric
statistical theory. However, in many cases the set of \(\lambda_{est}\)
spanned nearly the entire range of possible values (e.g., \(n=32\);
\(\lambda_{in}=0.5\): \(\lambda_{est}= 0.0\to 0.985\)), revealing that
estimates of \(\lambda\) were not a reliable indicator of input
phylogenetic signal. Importantly, the precision of \(\lambda_{est}\) was
not uniform across all levels of phylogenetic signal. The worst
precision was observed at intermediate levels of phylogenetic signal
(\(\lambda_{in}\approx0.5\)), while precision improved as input levels
approached the extremes of \(\lambda\)'s range (i.e.,
\(\lambda_{in}\rightarrow0\) \& \(\lambda_{in}\rightarrow1\)). Thus,
estimates of \(\lambda\) were least reflective of the true input signal
at intermediate values. Additionally, even at large levels of species
richness, we found that the range of \(\lambda_{est}\) still encompassed
a substantial portion of possible values (e.g., \(n=512\);
\(\lambda_{in}=0.5\): \(\lambda_{est} = 0.32\to 0.68\)). Likewise, the
same \(\lambda_{est}\) could be obtained from datasets containing vastly
different input levels of phylogenetic signal (e.g., \(n=512\);
\(\lambda_{est} = 0.5\); \(\lambda_{in} = 0.25 \to 0.65\)). These
findings were particularly unsettling when considered in light of our
literature survey. Over one quarter of the \(\lambda\) estimates
obtained in empirical studies (421 of 1,572) were between
\(\lambda=0.25\) and \(\lambda=0.75\) (Figure 1). This range reflected
the region that our simulations identified as being the least reliable
in terms of accurately characterizing levels of phylogenetic signal, yet
30\% of these mid-range empirical estimates were explicitly interpreted
in terms of the strength of phylogenetic signal that they represented
(i.e., weak, intermediate, strong phylogenetic signal). \hfill\break

Finally, when \(\lambda\) was co-estimated with regression parameters in
PGLS regression and ANOVA, the results of our simulations were quite
similar. Here, regression parameters (\(\beta\)) were accurately
estimated, confirming earlier findings of Revell (2010) (see Supporting
Information). However, estimates of phylogenetic signal (\(\lambda\))
were less precise (Figure 3; see also Supporting Information), and the
spread of \(\lambda_{est}\) was similar to that observed when
\(\lambda\) was estimated for only the dependent variable, as in Figure
2. Taken together, these findings reveal that \(\lambda_{est}\) does not
precisely characterize observed levels of phylogenetic signal in
phenotypic datasets, and that biological interpretations of the strength
of phylogenetic signal based on \(\lambda\) may be highly inaccurate.
\hfill\break

@@

Adams and Collyer (2019b) suggested that the strength of phylogenetic
signal could be represented as an effect size based on the \(Kappa\)
statistic and its empirical sampling distribution from permutation. Here
we formalize that suggestion, resulting in an effect size of:

\begin{align}
    Z_{K}=\frac{log(K_{obs})-\hat\mu_{log(K)}}{\hat\sigma_{log(K)}}
\end{align}

where \(K_{obs}\) is the observed phylogenetic signal, and \(\hat\mu_K\)
and \(\hat\sigma_K\) are the mean and standard deviation of the
empirical sampling distribution of \(log(Kappa)\) obtained via
permutation. Here the logarithm was used because \(Kappa\) takes only
positive values (\(0\to\infty\)) and its sampling distribution is
log-normally distributed (for a similar transformation when calculating
multivariate effect sizes see: Appendix 1 of Collyer and Adams 2018).
Similarly, an effect size based on \(\lambda\) could be envisioned as:

@@

In this case, \(\lambda_{obs}\) and \(\hat\sigma_\lambda\) are
empirically derived using maximum likelihood, as permutation approaches
have not been developed for evaluating \(\lambda\). Note also that under
the null hypothesis, no phylogenetic signal is expected (Freckleton et
al. 2002), and thus \(E(\lambda)=0\) under this condition. \hfill\break

To evaluate the utility of \(Z_K\) and \(Z_\lambda\) we calculated both
effect sizes for the simulated datasets generated above, and summarized
the precision of each using its variance (\(\sigma^2_{Z_K}\) and
\(\sigma^2_{Z_\lambda}\)). Results are found in Figure 4 (additional
results are found in the Supporting Information). Here two things are
evident. First, estimates of \(Z_K\) track the input phylogenetic signal
in a more linear fashion than do estimates of \(Z_\lambda\) (Figure
4A,B). Thus, actual changes in the strength of phylogenetic signal are
reflected more evenly in the corresponding values of the effect size
\(Z_K\). Second, the precision of \(Z_K\) is considerably more stable as
compared with \(Z_\lambda\). This may be seen by calculating the
coefficients of variation for the set of precision estimates (i.e.,
\(\sigma^2_{Z_K}\) and \(\sigma^2_{Z_\lambda}\)) across input levels of
phylogenetic signal. Here coefficients of variation in the precision of
\(Z_K\) were an order of magnitude smaller for than for \(Z_\lambda\)
(Figure 4C). This implied that estimates of the strength of phylogenetic
signal were more reliable and robust when using \(Z_K\) as compared with
\(Z_\lambda\). \hfill\break

@@

Once the magnitude of phylogenetic signal is characterized using
\(Z_K\), one may wish to compare such measures across datasets, to
determine whether the strength of phylogenetic signal is significantly
greater in one phenotypic trait as compared to another. As with other
effect sizes derived from permutation distributions (e.g., Adams and
Collyer 2016, 2019a), a two-sample test statistic may be calculated as:

@@

@@

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

It is common in comparative evolutionary studies to characterize the
phylogenetic signal in phenotypic traits to determine the extent to
which shared evolutionary history has generated trait covariation among
taxa. However, while numerous analytical approaches may be used to
quantify phylogenetic signal (e.g., Abouheif 1999; Gittleman and Kot
1990; Pagel 1999; Blomberg et al. 2003; Adams 2014a), methods that
explicitly measure the strength of phylogenetic signal, or facilitate
comparisons among datasets, have remained underdeveloped. In this study,
we evaluated the precision of one common measure, Pagel's \(\lambda\),
and explored its efficacy for characterizing the strength of
phylogenetic signal in phenotypic data. Using computer simulations, we
found that the precision of \(\lambda\) increased with increasing sample
sizes; a pattern noted previously (Boettiger et al. 2012), and one that
conformed with parametric statistical theory (Cohen 1988). However, we
also found that vastly different \(\lambda\) estimates could be obtained
from data containing the same level of phylogenetic signal, and that
similar \(\lambda\) estimates may be obtained from data containing
differing levels of phylogenetic signal. Further, the precision of
\(\lambda\) varied with the strength of phylogenetic signal, where lower
precision was observed when in data whose phylogenetic signal was of
intemediate strength. From these findings we conclude that \(\lambda\)
is not a reliable indicator of the observed strength of phylogenetic
signal in phenoytpic datasets, and that biological interpretations of
the strength of signal based on this parameter may innacurately
characterize such effects. \hfill\break

As an alternative, we described a standardized effect size (\(Z\)) for
assessing the strength of phylogenetic signal. \(Z\) expresses the
magnitude of phylogenetic signal as a standard normal deviate, which is
easily interpretable as the strength of phylogenetic signal relative to
the mean. We applied this concept to both \(\lambda\) and \(Kappa\), and
found that \(Z_K\) was a better estimate of the strength of phylogenetic
signal in phenotypic data. First, \(Z_K\) was more precise than
\(Z_{\lambda}\), and variation in this precision was more consistent
across the range of input levels of phylogenetic signal. Additionally,
values of \(Z_K\) more accurately tracked known levels of phylogenetic
signal, with changes in the actual strength of phylogenetic signal
reflected in a more linear fashion by concomite changes in the values of
\(Z_K\). Thus, \(Z_K\) holds promise as a measure of the relative
strength of phylogenetic signal that reflects the magnitude of this
effect in phenotypic data. We therefore recommend that future studies
interested in the strength of phylogenetic signal incorporate \(Z_K\) as
a statistical measure of this effect. \hfill\break

Based on the effect size \(Z_K\), we then proposed a two-sample test,
which provides a statistical means of determining whether the strength
of phylogenetic signal is greater in one phenotypic trait as compared to
another. Prior studies have summarized patterns of variation in
phylogenetic signal across datasets using summary test values, such as
\(Kappa\) (e.g., Blomberg et al. 2003). However, \(Kappa\) does not
scale linearly with input levels of phylogenetic signal, and its
variance increases (i.e., precision decreases) with increasing strength
of phylogenetic signal (Munkemuller et al. 2012; Diniz-Filho et al.
2012: see also Supporting Information). Thus, \(Kappa\) should not be
considered a standardized effect size that measures the strength of
phylogenetic signal on a common scale. By contrast, converting \(Kappa\)
to \(Z_K\) (via equation 2) alleviates these concerns, and facilitates
formal statistical comparisons of the strength of signal across
datasets. Thus when viewed from this perspective, the approach developed
here is statistically aligned with other statistical approaches such as
meta-analysis (sensu Hedges and Olkin 1985; Glass 1976; Arnqvist and
Wooster 1995), where summary statistics across datasets are converted to
standardized effect sizes for subsequent `higher order' statistical
summaries or comparisons. As such our approach enables evolutionary
biologists to quantitatively examine the relative strength of
phylogenetic signal across a wide range of phenotypic traits, and thus
opens the door for future discoveries that inform on how phenotypic
diversity accumulates in macroevolutionary time across the tree of life.
\hfill\break
